Predicting Success of Clinical Interviews via Sequential Analysis of Patient-Provider Interactions

### 2017-03-07 paper/AMIA/main.tex
first revision by Alex

\title{Sequence classifications to facilitate designing an effective intervention of motivational interview}

\author{Mehedi Hasan, BS$^{1}$, Alexander Kotov, PhD$^{1}$, April Carcone, PhD$^{2}$, Sylvie Naar-King, PhD$^{2}$}

@@

\noindent{\bf Abstract}
\textit{Childhood obesity is a serious public health concern in the United States and worldwide. Therefore, there is a need for identification of successful motivational interviews to facilitate the development of effective interventions for childhood obesity. This paper presents our work on the classification of utterance sequences into successful and unsuccessful behavior outcomes. Markov chain and hidden Markov models were employed and compared with a different order for the classification of utterance sequences. The classification performances achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using 1st order Markov chain and second order hidden Markov model, respectively. This shows that the behavior codes carry some inherited information to distinguish motivational interviews. We expect that the result will improve if we use a more discriminable data set. Furthermore, the results suggest us a new way to think about designing the motivational interview.}

\section*{Introduction}
In the recent times, more and more data from various domains being produced in the form of event sequences. Thus, sequence classification has become an important task in sequence mining. Sequences can contain discrete (e.g., symbolic sequences such as text, proteins, or DNA) or continuous events (e..g, time series such as ECGs or stocks), depending on the event types. Usually, sequences don't have explicit features and suffer from the high dimension of feature space. Even the sequential nature of features is too difficult to capture, which make sequence classification, a more challenging task than feature-based classifications. The sequence classification methods can be divided into three large categories: feature-based classification, distance based method, and model based classifications. In feature-based classification, a sequence is transformed into a single vector of features. A supervised learning algorithm such as a support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature} can be used to train the classifier. An improvement of feature-based classifier such as shapelet \cite{ye2009time} or pattern-based \cite{kudenko1998feature, lesh1999mining} approach is used for sequence classifications. On the other hand, distance based method measure the similarity between sequences to determine the quality of the classification significance. The most commonly used distance function is euclidian distance \cite{keogh2003need}, while Dynamic Time Wrapping \cite{keogh2000scaling} is used when more flexible matching is desired in time series data. The third category of sequence classification is model based classifiers such as hidden Markov model \cite{rabiner1989tutorial} (HMM) or other statistical models. Recently, a hierarchical approach \cite{nallam2016effective} is proposed for sequence classification.

Sequence classification has a broad range of real word applications such as information retrieval, genomic analysis, health informatics, finance and abnormal detection. In genomic research, sequence classification is widely used to detect the function of a new protein \cite{deshpande2002evaluation}. In health informatics, ECG can be used as a multi-dimensional time series data to classify an individual as healthy or patient with heart disease \cite{wei2006semi}. Another example is to classify rumor stance in Twitter \cite{lukasik2016hawkes}. In information retrieval, sequence classification is employed to build models for classifying documents into different topic categories \cite{sebastiani2002machine} such as sports, politics, news, finance and style. Sequence classification is also used in anomaly detection such as abnormal access to systems \cite{lane1999temporal}, malware detection \cite{drew2017polymorphic} and money laundering \cite{liu2008sequence}. Other interesting applications include classifying query log sequences to distinguish web-robots from human users \cite{duskin2009distinguishing, tan2004discovery}. In computer vision, sequence classification is applied for classifying images \cite{li2000image} and videos \cite{zhang2016tensor}.

In this paper, our target area of application is the motivational interview. In particular, this paper describes our work on the classification of utterance sequences. Classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment versus negative change talk or commitment is an important part of clinical research aimed at designing effective interventions for many conditions and disorders. In this study, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents (teens) and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. Therefore, there is a need for informatics-based methods to facilitate the development of effective interventions for childhood obesity. One approach to designing an effective intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change that can be sustained for as long as 34 months after an interview.

>>>>

\title{Predicting Success of Clinical Interviews via Sequential Analysis of Patient-Provider Interactions}

\author{Mehedi Hasan, BS$^{1}$, Alexander Kotov, PhD$^{1}$, April Carcone, PhD$^{2}$, Mind Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}

@@

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered sequences of past events to make predictions related to the observed sequence of events arises in many data analytics scenarios in healthcare informatics. In this paper, we focus on the analysis of patient-provider communication sequences in the context of clinical interviews and propose the methods for identifying the most effective communication sequences for triggering a particular type of patient response and dynamically predicting this type of response based on observed interactions. Such analysis is necessary to facilitate the development of effective interventions to address many public health conditions, such as childhood obesity. Our proposed probabilistic method for predicting a patient behavioral response in communication sequences can be utilized with both Markov chains or Hidden Markov Models. It achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using first and second order Markov chain and Hidden Markov Model, respectively. Our results indicate that individual short length patient-provider communication sequences are highly indicative of the progression and trajectory of clinical interviews and can be used to predict the overall success of motivational interviews. Our results have broad implications for qualitative public health research by providing a formal mechanism to facilitate the development of more effective behavioral interventions.}

\section*{Introduction}
Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics. The order of observations in a sequence is challenging to capture using features, which makes sequence classification a more challenging task than traditional classification. In general, sequence classification methods can be divided into three  categories: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. Distance-based methods measure the similarity between sequences to determine the quality of classification. The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. The third type of sequence classification methods, first create a probabilistic model of a sequence, such as Hidden Markov Model \cite{rabiner1989tutorial} (HMM).

Sequence classification has a broad range of real word applications from genomics and health informatics to finance and abnormality detection. In genomics research, sequence classification is widely used to detect the function of new proteins \cite{deshpande2002evaluation}. In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease \cite{wei2006semi}. Sequence classification is also used for anomaly detection such as abnormal access to systems \cite{lane1999temporal} and malware detection \cite{drew2017polymorphic}.

In this paper, we direct our focus towards patient-provider communication sequences in the context of clinical interviews. In particular, we propose a method to identify the most effective communication sequences for eliciting a specific type of patient response and . Classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment versus negative change talk or commitment is an important part of clinical research aimed at designing effective interventions for many behavioral conditions and disorders. In this study, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents (teens) and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. Therefore, there is a need for informatics-based methods to facilitate the development of effective interventions for childhood obesity. One approach to designing an effective intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change that can be sustained for as long as 34 months after an interview.
